A Comparison of R Tools for Nonlinear Least Squares Modeling

### 2024-02-08 copyedit240208/20230-15-copyeditresponse.Rmd
first pass complete

## Principal messages

Our work aimed at unifying nonlinear modeling functionality in R,
<!-- JN: removed "was" as suggesgted -->
ideally in a refactored `nls()` function.

For R **users**, we would advise adapting working scripts, preferably those
with documentation and using recent tools. If there is a suspicion that there
may be ill-conditioning, package `nlsr` or the example we give in the 
section "Comparison notes for formula-setup solutions" below of how to
find singular values of the Jacobian allow these diagnostics to be calculated.
<!-- JN: reference to section added ??have I done this correctly? -->

For R **developers**, we invite and encourage discussion of the design choices, 
since these
have downstream implications for ease of use, adaptation to new features, and 
efficiency of ongoing maintenance.
<!-- JN: Could not understand how copyedit "such as" would be applied -->

@@

`onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
is the vertical difference between the dependent variable and the functional
model. `onls` minimizes the sum of **orthogonal** residuals. The objective is
therefore different and involves nontrivial extra calculation. A vignette with
the package and the blog article 
https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
give some description with illustrative graphs. `onls` appears to be limited to 
problems with one independent and one dependent variable. The Wikipedia article
https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
ideas, with references to the literature. The approach needs a wider 
discussion and tutorial examples to allow its merits to be judged than can be
included here.

@@

The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth-curve modeling 
task that seems straightforward but is quite nasty. Its very succinct statement 
provides the "short reproducible example" much requested on R mailing lists. 
This is a real problem from a field researcher. The data and graph follow.

@@

Indeed, the argument only gets used when `nls()` is about to return its result
object, and the
element `model` is NULL unless the calling argument `model` is TRUE. (Using the same 
name could
be confusing.) Despite this, the model frame is used within the function code in 
the form of the object
`mf`. We feel that users could benefit from more extensive documentation and 
examples of its use since it is used to implement features like `subset`.

@@

The function `resid()` (an alias for `residuals()`) gives WEIGHTED residuals, 
as does, for example, `result$m$resid()`. The function `nlsModel()` 
allows us to compute residuals for particular coefficient sets.
We have had to extract `nlsModel()` from the base R code and include it via a code chunk 
(not echoed here for space) because it is not exported to the working namespace,
We could also explicitly `source()` this code.

@@

Hans Werner Borchers was helpful in developing the GSoC project motivating this article
and in comments on this and related work. Heather Turner co-mentored the project 
and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
clarify aspects of `selfStart` models and instigated the "Introduction to nlsr" 
vignette. Colin Gillespie (package `benchmarkme`) has been helpful in guiding our 
attempts to succinctly summarize computing environments. We thank one of the referees 
for pointing out that it is important to use background knowledge and graphs to
guide modeling.

# References

>>>>

<!-- ## Principal messages -->
<!-- Ed: How does this relate to the comparisons? -->
<!-- JN: It is a precis of the content.  -->

Our work aimed at unifying nonlinear modeling functionality in R,
<!-- JN: removed "was" as suggesgted -->
ideally in a refactored `nls()` function. The primary messages from this work
are:

- For R **users**, we would advise that it is most efficient to carry out
nonlinear modeling or least squares by adapting working scripts, preferably those
<!-- JN: added wording to clarify what we are advising users to do -->
with documentation and using recent tools. If there is a suspicion that there
may be ill-conditioning, package `nlsr` or the example we give in the 
section "Comparison notes for formula-setup solutions" below of how to
find singular values of the Jacobian allow these diagnostics to be calculated.
<!-- JN: reference to section added ??have I done this correctly? -->

- For R **developers**, we invite and encourage discussion of the design choices, 
since these
have downstream implications for ease of use, adaptation to new features, and 
efficiency of ongoing maintenance.
<!-- JN: Could not understand how copyedit "such as" would be applied -->

@@

<!-- JN: copyedit suggestion incorporated -->
`onls` (@R-onls-manual) is used for optimising and estimating nonlinear models 
by minimizing the sum of squares of **orthogonal** residuals rather than vertical
residuals. The objective is therefore different and involves nontrivial extra
calculation. A vignette with the package and the blog article 
https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
give some description with illustrative graphs. `onls` appears to be limited to 
problems with one independent and one dependent variable. The Wikipedia article
https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
<!-- Ed. Anything specific from the Wikipedia article that is worth citing? -->
<!-- JN: Commentary would add too much material of tangential nature. -->
ideas, with references to the literature. The approach needs a wider 
discussion and tutorial examples to allow its merits to be judged than can be
included here.

@@

The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth-curve modeling 
task. Its very succinct statement 
provides the "short reproducible example" much requested on R mailing lists. 
The problem is small and seemingly straightforward, yet presents such 
difficulties that optimization researchers have asked if it is contrived
rather than a real problem from a field researcher. The data and graph follow.
<!-- JN: merged copyedit comments to explain "nasty". -->

@@

Indeed, the argument only gets used when `nls()` is about to return its result
object, and the
element `model` is NULL unless the calling argument `model` is TRUE. Using the 
same name for both function argument and object element could be confusing.
<!-- JN: reworded to make clear our concern. -->
Despite this, the model frame is used within the function code in 
the form of the object
`mf`. We feel that users could benefit from more extensive documentation and 
examples of its use since it is used to implement features like `subset`.

@@

The function `resid()` (an alias for `residuals()`) gives weighted residuals
like `result$m$resid()`. The function `nlsModel()` 
<!-- JN: copyedit suggestion accepted -->
allows us to compute residuals for particular coefficient sets.
We have had to extract `nlsModel()` from the base R code and include it via a code chunk 
(not echoed here for space) because it is not exported to the working namespace,
We could also explicitly `source()` this code.

@@

Hans Werner Borchers was helpful in developing the GSoC project motivating this article
and in comments on this and related work. Heather Turner co-mentored the project 
and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
clarify aspects of `selfStart` models and instigated the "Introduction to nlsr" 
vignette. Colin Gillespie (package `benchmarkme`) has been helpful in guiding our 
attempts to succinctly summarize computing environments. We thank one of the referees 
for pointing out that it is important to use background knowledge and graphs to
guide modeling. Simon Urbanek gave some important practical help with preparing
files for publication.
<!-- JN: added Simon. Removed # References which duplicated subhead -->

### 2024-02-08 copyedit240208/20230-15-copyeditresponse.tex
first pass complete

\hypertarget{principal-messages}{%
\subsection{Principal messages}\label{principal-messages}}

Our work aimed at unifying nonlinear modeling functionality in R,
ideally in a refactored \texttt{nls()} function.

For R \textbf{users}, we would advise adapting working scripts, preferably those
with documentation and using recent tools. If there is a suspicion that there
may be ill-conditioning, package \texttt{nlsr} or the example we give in the
section ``Comparison notes for formula-setup solutions'' below of how to
find singular values of the Jacobian allow these diagnostics to be calculated.

For R \textbf{developers}, we invite and encourage discussion of the design choices,
since these
have downstream implications for ease of use, adaptation to new features, and
efficiency of ongoing maintenance.

@@

\texttt{onls} (Spiess, Andrej-Nikolai (2022)) -- The usual optimization in estimating nonlinear models
is the vertical difference between the dependent variable and the functional
model. \texttt{onls} minimizes the sum of \textbf{orthogonal} residuals. The objective is
therefore different and involves nontrivial extra calculation. A vignette with
the package and the blog article
\url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
give some description with illustrative graphs. \texttt{onls} appears to be limited to
problems with one independent and one dependent variable. The Wikipedia article
\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an overview of some
ideas, with references to the literature. The approach needs a wider
discussion and tutorial examples to allow its merits to be judged than can be
included here.

@@

The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth-curve modeling
task that seems straightforward but is quite nasty. Its very succinct statement
provides the ``short reproducible example'' much requested on R mailing lists.
This is a real problem from a field researcher. The data and graph follow.

@@

Indeed, the argument only gets used when \texttt{nls()} is about to return its result
object, and the
element \texttt{model} is NULL unless the calling argument \texttt{model} is TRUE. (Using the same
name could
be confusing.) Despite this, the model frame is used within the function code in
the form of the object
\texttt{mf}. We feel that users could benefit from more extensive documentation and
examples of its use since it is used to implement features like \texttt{subset}.

@@

The function \texttt{resid()} (an alias for \texttt{residuals()}) gives WEIGHTED residuals,
as does, for example, \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}
allows us to compute residuals for particular coefficient sets.
We have had to extract \texttt{nlsModel()} from the base R code and include it via a code chunk
(not echoed here for space) because it is not exported to the working namespace,
We could also explicitly \texttt{source()} this code.

@@

Hans Werner Borchers was helpful in developing the GSoC project motivating this article
and in comments on this and related work. Heather Turner co-mentored the project
and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
clarify aspects of \texttt{selfStart} models and instigated the ``Introduction to nlsr''
vignette. Colin Gillespie (package \texttt{benchmarkme}) has been helpful in guiding our
attempts to succinctly summarize computing environments. We thank one of the referees
for pointing out that it is important to use background knowledge and graphs to
guide modeling.

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{references-1}{%
\section*{References}\label{references-1}}
\addcontentsline{toc}{section}{References}

>>>>

Our work aimed at unifying nonlinear modeling functionality in R,
ideally in a refactored \texttt{nls()} function. The primary messages from this work
are:

\begin{itemize}
\item
  For R \textbf{users}, we would advise that it is most efficient to carry out
  nonlinear modeling or least squares by adapting working scripts, preferably those
  with documentation and using recent tools. If there is a suspicion that there
  may be ill-conditioning, package \texttt{nlsr} or the example we give in the
  section ``Comparison notes for formula-setup solutions'' below of how to
  find singular values of the Jacobian allow these diagnostics to be calculated.
\item
  For R \textbf{developers}, we invite and encourage discussion of the design choices,
  since these
  have downstream implications for ease of use, adaptation to new features, and
  efficiency of ongoing maintenance.
\end{itemize}

@@

\texttt{onls} (Spiess, Andrej-Nikolai (2022)) is used for optimising and estimating nonlinear models
by minimizing the sum of squares of \textbf{orthogonal} residuals rather than vertical
residuals. The objective is therefore different and involves nontrivial extra
calculation. A vignette with the package and the blog article
\url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
give some description with illustrative graphs. \texttt{onls} appears to be limited to
problems with one independent and one dependent variable. The Wikipedia article
\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an overview of some
ideas, with references to the literature. The approach needs a wider
discussion and tutorial examples to allow its merits to be judged than can be
included here.

@@

The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth-curve modeling
task. Its very succinct statement
provides the ``short reproducible example'' much requested on R mailing lists.
The problem is small and seemingly straightforward, yet presents such
difficulties that optimization researchers have asked if it is contrived
rather than a real problem from a field researcher. The data and graph follow.

@@

Indeed, the argument only gets used when \texttt{nls()} is about to return its result
object, and the
element \texttt{model} is NULL unless the calling argument \texttt{model} is TRUE. Using the
same name for both function argument and object element could be confusing.
Despite this, the model frame is used within the function code in
the form of the object
\texttt{mf}. We feel that users could benefit from more extensive documentation and
examples of its use since it is used to implement features like \texttt{subset}.

@@

The function \texttt{resid()} (an alias for \texttt{residuals()}) gives weighted residuals
like \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}
allows us to compute residuals for particular coefficient sets.
We have had to extract \texttt{nlsModel()} from the base R code and include it via a code chunk
(not echoed here for space) because it is not exported to the working namespace,
We could also explicitly \texttt{source()} this code.

@@

Hans Werner Borchers was helpful in developing the GSoC project motivating this article
and in comments on this and related work. Heather Turner co-mentored the project
and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
clarify aspects of \texttt{selfStart} models and instigated the ``Introduction to nlsr''
vignette. Colin Gillespie (package \texttt{benchmarkme}) has been helpful in guiding our
attempts to succinctly summarize computing environments. We thank one of the referees
for pointing out that it is important to use background knowledge and graphs to
guide modeling. Simon Urbanek gave some important practical help with preparing
files for publication.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}
